---
title: "Reproducible Research - Course Project 1"
author: "Kun He"
date: "Friday, September 2, 2016"
---


#Introduction

This document is the first project for the **Reproducible Research** course in Coursera's Data Science specialization track. This document is written in a single R markdown document that can be processed by knitr and transformed into an HTML file.

The data used in this project is [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip), data about personal movement using activity monitoring devices such as a [Fitbit](http://www.fitbit.com/nl), [Nike Fuelband](http://www.nike.com/us/en_us/c/nikeplus-fuelband), or [Jawbone Up](https://jawbone.com/up). This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.


## R Preparations

Before the analysis, first prepare the basic settings and activate relevant libraries for the analysis.
Also remember to set the dirrectory using setwd function.
```{r setup}
library(knitr)
opts_chunk$set(echo = TRUE)

library(lattice)
library(ggplot2)
library(dplyr)
```

## 1. Reading and preprocessing the data

Download the dataset [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) into the working dirrectory, unzip and load data into data frame `data`. Then use read.csv() function to load the data into data frame `data`.
```{r}
unzip(zipfile="activity.zip")
data <- read.csv("activity.csv")
```

## 2. Histogram of the total number of steps taken each day

Create a histogram showing the distribution of steps taken.
```{r} 
total.steps <- tapply(data$steps, data$date, FUN=sum, na.rm=TRUE)
qplot(total.steps, binwidth=1000, main = paste("Total Steps Each Day (missing data ignored)"), xlab="total number of steps taken each day")
```

## 3. Mean and median number of steps taken each day

```{r} 
rmean <- mean(total.steps, na.rm=TRUE)
rmedian <- median(total.steps, na.rm=TRUE)
```
The `mean` is `r rmean` and the `median` is `r rmedian`.


## 4. Time series plot of the average number of steps taken

Create a time series plot by summarizing the average steps taken (y-axis) based on a 5-minute interval (x-axis).

```{r} 
averages <- aggregate(x=list(steps=data$steps), by=list(interval=data$interval),
                      FUN=mean, na.rm=TRUE)
ggplot(data=averages, aes(x=interval, y=steps)) +
    geom_line() +
    xlab("5-minute interval") +
    ylab("average number of steps taken")

```

## 5. The 5-minute interval that, on average, contains the maximum number of steps

Check which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps
```{r}
averages[which.max(averages$steps),]
```

## 6. Code to describe and show a strategy for imputing missing data

#### Check the missing data status

```{r} 
missing <- is.na(data$steps)
table(missing)
```

#### Replace each missing value with the mean value of its 5-minute interval

```{r} 
fill.value <- function(steps, interval) {
    filled <- NA
    if (!is.na(steps))
        filled <- c(steps)
    else
        filled <- (averages[averages$interval==interval, "steps"])
    return(filled)
}
filled.data <- data
filled.data$steps <- mapply(fill.value, filled.data$steps, filled.data$interval)
```

## 7. Histogram of the total number of steps taken each day after missing values are imputed

Create a new histogram showing the distribution of steps taken based on imputed data.
```{r} 
total.steps <- tapply(filled.data$steps, filled.data$date, FUN=sum)
qplot(total.steps, binwidth=1000, main = paste("Total Steps Each Day (missing data imputed)"), xlab="total number of steps taken each day")

#Mean and median number of steps taken each day
mean(total.steps)
median(total.steps)
```

## 8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

First we create a new factor variable `day` with two levels - "weekday" and "weekend" based on `date` variable.

```{r} 
weekday.or.weekend <- function(date) {
    day <- weekdays(date)
    if (day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
        return("weekday")
    else if (day %in% c("Saturday", "Sunday"))
        return("weekend")
    else
        stop("invalid date")
}
filled.data$date <- as.Date(filled.data$date)
filled.data$day <- sapply(filled.data$date, FUN=weekday.or.weekend)
```

Create a pair time series plots for both weekday and weekend scenarios.

```{r} 
averages <- aggregate(steps ~ interval + day, data=filled.data, mean)
ggplot(averages, aes(interval, steps)) + geom_line() + facet_grid(day ~ .) +
    xlab("5-minute interval") + ylab("Number of steps")
```